Guilt, shame, and anti-war action in an authoritarian country at war

### 2023-04-27 manuscript.tex
Sorted supplementary tables except S1

@@

@@

\noindent All study materials, data, and code can be found on the OSF platform: \url{https://osf.io/4pd2v/?view_only=23597fdc19424706818a72fc0df43009}

@@

This prediction that guilt would motivate reconciliatory behavior, whereas shame would motivate avoidance is not well supported by the empirical evidence available so far. A recent meta-analysis of group-based emotions ((Hakim et al., 2021), 101 effect sizes, 58 samples, N = 10,305) found that guilt, shame, and anger are equally predictive of support for reparations, with a strong average effect size of r = .44 and no significant differences between emotions. Furthermore, the distinction between moral shame and image shame (Allpress et al., 2010; Rees et al., 2013) helps explain which elements of shame are predictive of reconciliatory behaviors and which are predictive of avoidant behaviors. This distinction is made based on different appraisals that the feeling of shame can be accompanied with: we feel moral shame when the group members' behavior violates important moral standards or values of the group, and we feel image shame when the group members' behavior tarnishes the ingroup's reputation (Rees et al., 2013). Studies that make the distinction between moral shame and image shame find that moral shame is predictive of reconciliatory behaviors on par or more strongly than guilt, whereas image shame is predictive of distancing and avoidance (Allpress et al., 2014; Allpress et al., 2010; Grigoryan and Efremova, 2017; Nooitgedagt et al., 2021; Rees et al., 2013). We therefore expect that moral shame and guilt, but not image shame, will predict a stronger intention to act against the war in Russia (H1).

@@

We collected a quota sample, representative of the Russian population by gender and age. Participants were recruited via the online crowdsourcing platform Yandex Toloka in August 2022. This was the only feasible option to collect fully anonymized data in Russia, which was crucial, given that the repercussions for expressing opposition to the war can be severe. Participants received \$ 0.5 (30 RUB) for their participation. A total of N=1011 participants gave full informed consent to participate in the study. Thirty participants were excluded since they dropped out of the study before completion, and another eight were excluded for failing more than one out of three attention checks. The effective sample size is N=973.

@@

\textbf{\emph{Anti-war actions}}. We measured past behavior by asking how many of the eight political actions listed did the participant do to oppose the war in Ukraine (e.g., ``signing a petition'', ``participating in a demonstration'', etc.). The responses ranged from 0 to 8, with 82\% of participants reporting zero actions. Behavioral intention to participate in any political action against the war was measured with 3 items (``I want to/intend to/plan to take part in political action against the war in Ukraine''), on a 5-point scale from 1 -- ` Completely disagree' to 5 -- `Completely agree' (\(\alpha\) = .95; 73\% of participants had a mean score of 1). Finally, we asked about the probability that the person will take any political action, from 1 -- highly improbable, 10 -- (almost) certainly (69\% answered `1').

\textbf{\emph{Attitude towards the war}} was measured with 3 items, on a bipolar 5-pt Osgood-style scale: ``The special military operation in Ukraine is\ldots{}'': bad-good, harmful-beneficial, useless-useful (\(\alpha\) = .90, M = 2.88, SD = 1.29).

@@

\emph{Support for democratic values} was measured with a 6-item scale from Canetti-Nisim and Beit-Hallahmi (2007). Example item: ``Every citizen has the right to take his convictions to the street if necessary'' (\(\alpha\) = .69).

\emph{Political alienation} was measured with four items from Olsen (1969). Example item: ``There is not much that people like me can do to influence actions of the government'' (\(\alpha\) = .73).

\emph{Political cynicism} was measured with eight items from Pattyn et al. (2012). Example item: ``Politicians pretend to care more about people than they really do'' (\(\alpha\) = .91).

\emph{Political responsibility.} We formulated five items aimed at capturing participants' beliefs about political responsibility: ``The government will face difficulties if it tries to do something that most people in the country do not agree with'', ``Participation in political life is a responsibility of every person before themselves and their fellow countrymen'', ``If I see that the country is going in the wrong direction, it is my duty to get my voice heard by the government'', ``The citizens of Russia are accountable for the actions of their government'', ``If the majority of the citizens are against some decision made by the government, they can influence that decision'' (\(\alpha\) = .71).

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = 0.803, RMSEA = 0.128, SRMR = 0.11). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = 0.947, RMSEA = 0.068, SRMR = 0.068). This modified model includes the political responsibility factor with 5 items, political alienation factor with 3 items (the item ``I believe public officials don't care much what people like me think'' was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: ``politicians are evil'' (4 items, e.g., ``Politicians are only interested in getting and maintaining power'') and ``politics is evil'' (4 items, e.g., ``No man can hope to stay honest once he enters politics'').

\textbf{\emph{Positive and negative emotions.}} We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale (Watson et al., 1988). Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-pt scale (1 -- `do not experience at all' to 5 -- `extremely'). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt and/or shame, respectively (we took this precaution because of negative prior experiences when using these measures in Russia; see Grigoryan et al. (2018); Grigoryan and Efremova (2017), for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame) using CFA and found the fit to be unsatisfactory (CFI = 0.862, RMSEA = 0.146, SRMR = 0.08). A 3-factor structure with positive emotions (joy, hope, enthusiasm), negative dominant emotions (anger, disgust, contempt) and negative submissive emotions (fear, sadness, depression) showed a satisfactory fit to the data (CFI = 0.916, RMSEA = 0.117, SRMR = 0.066) (see Mehrabian, 1980), for the distinction between dominant and submissive emotions).

\textbf{\emph{Group-based guilt and shame.}} We used the measures of group-based guilt, image shame, and moral shame developed by Rees et al. (2013), Russian adaptation by Grigoryan and Efremova (2017). All items are answered on a 5-pt scale from 1 -- `Completely disagree' to 5 -- `Completely agree'. Guilt was measured with 3 items (e.g., ``I feel guilty for the manner in which Ukrainian people have been treated by Russians'', \(\alpha\) = .89), image shame with 5 items (``I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine'', \(\alpha\) = .94), and moral shame with 4 items (``I feel ashamed because Russia's actions with regard to Ukraine have been immoral'', \(\alpha\) = .93). All missing values were replaced with zeros, since participants who didn't see the scale were the ones who reported no feelings of guilt or shame in the PANAS.

@@

In a series of linear regressions, we tested whether political beliefs predict experiences of group-based shame and guilt. As Table 1 indicates, participants with stronger democratic values and those with stronger belief that politicians are evil reported stronger experiences of group-based guilt, moral shame, and image shame. The effect size of democratic values was at least twice as strong as that of political cynicism (\(\beta\) \(\approx\) .40 vs.~\(\beta\) \(\approx\) .15). Political responsibility beliefs also predicted guilt, albeit weakly. We ran a path model to account for correlations between the different political beliefs and between the three emotions (Table S1, SM). The results were essentially identical. Overall, political beliefs explained 22-26\% of variance in experiences of group-based guilt and shame and the endorsement of democratic values was the main driver of these experiences. The results of the zero-inflated models (Table S2, SM) were essentially identical when predicting strength of the emotions. Only democratic values predicted the occurrence of emotions (zero vs.~non-zero scores).

\hypertarget{table-1}{%
\subsection*{Table 1}\label{table-1}}
\addcontentsline{toc}{subsection}{Table 1}

Predicting group-based guilt and shame from political beliefs

\begin{table}[H]
\centering\begingroup\fontsize{8}{10}\selectfont

\begin{tabular}{llccccccccccc}
\toprule
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Moral Shame} & \multicolumn{4}{c}{Image Shame} & \multicolumn{4}{c}{Guilt} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13}
  & $\beta$ & $b$ & $SE$ & $p$ & $\beta$ & $b$ & $SE$ & $p$ & $\beta$ & $b$ & $SE$ & $p$\\
\midrule
Intercept & 0 & -2.89 & 0.32 & <0.001 & 0 & -2.8500 & 0.34 & <0.001 & 0 & -2.930 & 0.34 & <0.001\\
Democratic Values & 0.41 & 0.64 & 0.05 & <0.001 & 0.4 & 0.6400 & 0.05 & <0.001 & 0.41 & 0.650 & 0.05 & <0.001\\
Political Cynicism (politicians) & 0.18 & 0.20 & 0.05 & <0.001 & 0.16 & 0.1800 & 0.05 & <0.001 & 0.09 & 0.100 & 0.05 & 0.039\\
Political Cynicism (politics) & -0.05 & -0.05 & 0.04 & 0.189 & -0.03 & -0.0300 & 0.04 & 0.458 & -0.03 & -0.030 & 0.04 & 0.455\\
Political Alienation & -0.05 & -0.06 & 0.04 & 0.122 & -0.03 & -0.0400 & 0.04 & 0.380 & 0.01 & 0.020 & 0.04 & 0.663\\
\addlinespace
Political Responsibility & 0.04 & 0.05 & 0.04 & 0.190 & 0.03 & 0.0400 & 0.04 & 0.347 & 0.07 & 0.100 & 0.04 & 0.022\\
\midrule
R$^{2}$ &  & 0.26 &  &  &  & 0.2345 &  &  &  & 0.221 &  & \\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

@@

In a series of linear regressions, we tested whether guilt, image shame, and moral shame are predictive of anti-war actions in the past and in the future. Only moral shame was consistently related to anti-war action and the results were consistent in linear (Table 2) and zero-inflated negative binomial models (Table S3, SM). Group-based guilt and shame explained about 18\% of variance in past behavior and about 40\% of variance in future behavior.

\hypertarget{table-2}{%
\subsection*{Table 2}\label{table-2}}
\addcontentsline{toc}{subsection}{Table 2}

Predicting anti-war action from group-based guilt and shame

\begin{table}[H]
\centering\begingroup\fontsize{8}{10}\selectfont

\begin{tabular}{lllllllllllll}
\toprule
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Past Behaviour} & \multicolumn{4}{c}{Behavioural Intention} & \multicolumn{4}{c}{Action Probability} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13}
  & $\beta$ & b & SE & p & $\beta$ & b & SE & p & $\beta$ & b & SE & p\\
\midrule
Intercept & 0 & 0.1 & 0.04 & 0.0094 & 0 & 1.09 & 0.03 & <0.001 & 0 & 1.21 & 0.07 & <0.001\\
Guilt & 0.02 & 0.01 & 0.04 & 0.7562 & 0 & 0 & 0.03 & 0.942 & 0.05 & 0.06 & 0.06 & 0.326\\
Image Shame & -0.07 & -0.05 & 0.05 & 0.3926 & 0.1 & 0.06 & 0.04 & 0.154 & -0.02 & -0.02 & 0.1 & 0.826\\
Moral Shame & 0.48 & 0.31 & 0.06 & <0.001 & 0.56 & 0.33 & 0.05 & <0.001 & 0.59 & 0.77 & 0.11 & <0.001\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

@@

Behavioral intentions and action probability correlated at r = 0.86, p \textless0.001, so we scaled and combined the two measures into a single indicator of future anti-war action. We ran a linear regression predicting future anti-war behavior from attitude towards the war, negative and positive emotions, and shame and guilt. Table 3 presents the results. Overall, the model explained 47\% of variance in future behavior. Anti-war behavioral intentions were best predicted by moral shame (\(\beta\) = 0.36, p \textless0.001), followed by negative dominant emotions (\(\beta\) = 0.24, p \textless0.001), and only then attitude (\(\beta\) = -0.15, p \textless0.001). Negative submissive emotions had a small negative effect on intentions to act (\(\beta\) = -0.08, p = 0.015).

\hypertarget{table-3}{%
\subsection*{Table 3}\label{table-3}}
\addcontentsline{toc}{subsection}{Table 3}

Predicting future behavior from emotions and attitudes

\begin{table}[H]
\centering\begingroup\fontsize{8}{10}\selectfont

\begin{tabular}{llllll}
\toprule
  & $\beta$ & b & SE & t-value & p\\
\midrule
Intercept & 0 & -0.1 & 0.12 & -0.82 & 0.410\\
Emotions: Positive & -0.03 & -0.03 & 0.03 & -0.96 & 0.337\\
Emotions: Negative Dominant & 0.24 & 0.19 & 0.03 & 6.4 & <0.001\\
Emotions: Negative Submissive & -0.08 & -0.07 & 0.03 & -2.43 & 0.015\\
Attitudes towards the war & -0.15 & -0.12 & 0.03 & -3.77 & <0.001\\
\addlinespace
Group-based moral shame & 0.36 & 0.22 & 0.05 & 4.63 & <0.001\\
Group-based image shame & 0 & 0 & 0.04 & 0.03 & 0.977\\
Group-based guilt & 0.03 & 0.02 & 0.03 & 0.76 & 0.450\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

The results of the zero-inflated negative binomial model (Table S4, SM) suggested that different predictors account for the presence vs.~absence of intentions to act versus the strength of these intentions. Only moral shame (b = -.36, p = .036) and attitude (b = .38, p = .001) predicted the probability of no action, whereas only negative emotions predicted the strength of these intentions. Negative dominant emotions predicted stronger intentions to act (b = .24, p \textless{} .001), whereas negative submissive emotions predicted weaker intentions to act (b = -.14, p = .013).

@@

>>>>

\usepackage{fancyhdr}
\pagestyle{fancy}

@@

\fancyhead[LH]{GUILT, SHAME, AND ANTI-WAR ACTION}

@@

\noindent All study materials, data, and code can be found on the OSF platform:

\noindent \url{https://osf.io/4pd2v/?view_only=23597fdc19424706818a72fc0df43009}

@@

This prediction that guilt would motivate reconciliatory behavior, whereas shame would motivate avoidance is not well supported by the empirical evidence available so far. A recent meta-analysis of group-based emotions ((Hakim et al., 2021), 101 effect sizes, 58 samples, N = 10,305) found that guilt, shame, and anger are equally predictive of support for reparations, with a strong average effect size of \emph{r} = .44 and no significant differences between emotions. Furthermore, the distinction between moral shame and image shame (Allpress et al., 2010; Rees et al., 2013) helps explain which elements of shame are predictive of reconciliatory behaviors and which are predictive of avoidant behaviors. This distinction is made based on different appraisals that the feeling of shame can be accompanied with: we feel moral shame when the group members' behavior violates important moral standards or values of the group, and we feel image shame when the group members' behavior tarnishes the ingroup's reputation (Rees et al., 2013). Studies that make the distinction between moral shame and image shame find that moral shame is predictive of reconciliatory behaviors on par or more strongly than guilt, whereas image shame is predictive of distancing and avoidance (Allpress et al., 2014; Allpress et al., 2010; Grigoryan and Efremova, 2017; Nooitgedagt et al., 2021; Rees et al., 2013). We therefore expect that moral shame and guilt, but not image shame, will predict a stronger intention to act against the war in Russia (H1).

@@

We collected a quota sample, representative of the Russian population by gender and age. Participants were recruited via the online crowdsourcing platform Yandex Toloka in August 2022. This was the only feasible option to collect fully anonymized data in Russia, which was crucial, given that the repercussions for expressing opposition to the war can be severe. Participants received \$0.5 (30 RUB) for their participation. A total of N=1011 participants gave full informed consent to participate in the study. Thirty participants were excluded since they dropped out of the study before completion, and another eight were excluded for failing more than one out of three attention checks. The effective sample size is N=973.

@@

\textbf{\emph{Anti-war actions}}. We measured \emph{past behavior} by asking how many of the eight political actions listed did the participant do to oppose the war in Ukraine (e.g., ``signing a petition'', ``participating in a demonstration'', etc.). The responses ranged from 0 to 8, with 82\% of participants reporting zero actions. \emph{Behavioral intention} to participate in any political action against the war was measured with 3 items (``I want to/intend to/plan to take part in political action against the war in Ukraine''), on a 5-point scale from 1 -- ` Completely disagree' to 5 -- `Completely agree' (\(\alpha\) = 0.95; 73\% of participants had a mean score of 1). Finally, we asked about \emph{the probability that the person will take any political action}, from 1 -- highly improbable, 10 -- (almost) certainly (69\% answered `1').

\textbf{\emph{Attitude towards the war}} was measured with 3 items, on a bipolar 5-pt Osgood-style scale: ``The special military operation in Ukraine is\ldots{}'': bad-good, harmful-beneficial, useless-useful (\(\alpha\) = 0.9, M = 2.88, SD = 1.29).

@@

\emph{Support for democratic values} was measured with a 6-item scale from Canetti-Nisim and Beit-Hallahmi (2007). Example item: ``Every citizen has the right to take his convictions to the street if necessary'' (\(\alpha\) = 0.69).

\emph{Political alienation} was measured with four items from Olsen (1969). Example item: ``There is not much that people like me can do to influence actions of the government'' (\(\alpha\) = 0.73).

\emph{Political cynicism} was measured with eight items from Pattyn et al. (2012). Example item: ``Politicians pretend to care more about people than they really do'' (\(\alpha\) = 0.91).

\emph{Political responsibility.} We formulated five items aimed at capturing participants' beliefs about political responsibility: ``The government will face difficulties if it tries to do something that most people in the country do not agree with'', ``Participation in political life is a responsibility of every person before themselves and their fellow countrymen'', ``If I see that the country is going in the wrong direction, it is my duty to get my voice heard by the government'', ``The citizens of Russia are accountable for the actions of their government'', ``If the majority of the citizens are against some decision made by the government, they can influence that decision'' (\(\alpha\) = 0.71).

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = 0.803, RMSEA = 0.128, SRMR = 0.11). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = 0.947, RMSEA = 0.068, SRMR = 0.068). This modified model includes the \emph{political responsibility} factor with 5 items, \emph{political alienation} factor with 3 items (the item ``I believe public officials don't care much what people like me think'' was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: ``\emph{politicians are evil}'' (4 items, e.g., ``Politicians are only interested in getting and maintaining power'') and ``\emph{politics is evil}'' (4 items, e.g., ``No man can hope to stay honest once he enters politics'').

\textbf{\emph{Positive and negative emotions.}} We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale (Watson et al., 1988). Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-pt scale (1 -- `do not experience at all' to 5 -- `extremely'). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt and/or shame, respectively (we took this precaution because of negative prior experiences when using these measures in Russia; see Grigoryan et al. (2018); Grigoryan and Efremova (2017), for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame) using CFA and found the fit to be unsatisfactory (CFI = 0.862, RMSEA = 0.146, SRMR = 0.08). A 3-factor structure with positive emotions (joy, hope, enthusiasm), negative dominant emotions (anger, disgust, contempt) and negative submissive emotions (fear, sadness, depression) showed a satisfactory fit to the data (CFI = 0.916, RMSEA = 0.117, SRMR = 0.066) (see Mehrabian (1980), for the distinction between dominant and submissive emotions).

\textbf{\emph{Group-based guilt and shame.}} We used the measures of group-based guilt, image shame, and moral shame developed by Rees et al. (2013), Russian adaptation by Grigoryan and Efremova (2017). All items are answered on a 5-pt scale from 1 -- `Completely disagree' to 5 -- `Completely agree'. Guilt was measured with 3 items (e.g., ``I feel guilty for the manner in which Ukrainian people have been treated by Russians'', \(\alpha\) = 0.89), image shame with 5 items (``I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine'', \(\alpha\) = 0.94), and moral shame with 4 items (``I feel ashamed because Russia's actions with regard to Ukraine have been immoral'', \(\alpha\) = 0.93). All missing values were replaced with zeros, since participants who didn't see the scale were the ones who reported no feelings of guilt or shame in the PANAS.

@@

In a series of linear regressions, we tested whether political beliefs predict experiences of group-based shame and guilt. As Table \ref{tab:Table1} indicates, participants with stronger democratic values and those with stronger belief that politicians are evil reported stronger experiences of group-based guilt, moral shame, and image shame. The effect size of democratic values was at least twice as strong as that of political cynicism (\(\beta\) \(\approx\) .40 vs.~\(\beta\) \(\approx\) .15). Political responsibility beliefs also predicted guilt, albeit weakly. We ran a path model to account for correlations between the different political beliefs and between the three emotions (Table \ref{tab:TableS1}, SM). The results were essentially identical. Overall, political beliefs explained 22-26\% of variance in experiences of group-based guilt and shame and the endorsement of democratic values was the main driver of these experiences. The results of the zero-inflated models (Table \ref{tab:TableS1}, SM) were essentially identical when predicting strength of the emotions. Only democratic values predicted the occurrence of emotions (zero vs.~non-zero scores).

\begin{table}[H]

\caption{\label{tab:Table1}Predicting group-based guilt and shame from political beliefs}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llccccccccccc}
\toprule
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Moral Shame} & \multicolumn{4}{c}{Image Shame} & \multicolumn{4}{c}{Guilt} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13}
  & $\beta$ & $b$ & $SE$ & $p$ & $\beta$ & $b$ & $SE$ & $p$ & $\beta$ & $b$ & $SE$ & $p$\\
\midrule
Intercept & 0 & -2.89 & 0.32 & <0.001 & 0 & -2.850 & 0.34 & <0.001 & 0 & -2.930 & 0.34 & <0.001\\
Democratic Values & 0.41 & 0.64 & 0.05 & <0.001 & 0.4 & 0.640 & 0.05 & <0.001 & 0.41 & 0.650 & 0.05 & <0.001\\
Political Cynicism (politicians) & 0.18 & 0.20 & 0.05 & <0.001 & 0.16 & 0.180 & 0.05 & <0.001 & 0.09 & 0.100 & 0.05 & 0.039\\
Political Cynicism (politics) & -0.05 & -0.05 & 0.04 & 0.189 & -0.03 & -0.030 & 0.04 & 0.458 & -0.03 & -0.030 & 0.04 & 0.455\\
Political Alienation & -0.05 & -0.06 & 0.04 & 0.122 & -0.03 & -0.040 & 0.04 & 0.380 & 0.01 & 0.020 & 0.04 & 0.663\\
\addlinespace
Political Responsibility & 0.04 & 0.05 & 0.04 & 0.190 & 0.03 & 0.040 & 0.04 & 0.347 & 0.07 & 0.100 & 0.04 & 0.022\\
\midrule
R$^{2}$ &  & 0.26 &  &  &  & 0.234 &  &  &  & 0.221 &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

In a series of linear regressions, we tested whether guilt, image shame, and moral shame are predictive of anti-war actions in the past and in the future. Only moral shame was consistently related to anti-war action and the results were consistent in linear (Table \ref{tab:Table2}) and zero-inflated negative binomial models (Table \ref{tab:TableS3}, SM). Group-based guilt and shame explained about 18\% of variance in past behavior and about 40\% of variance in future behavior.

\begin{table}[H]

\caption{\label{tab:Table2}Predicting anti-war action from group-based guilt and shame}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lllllllllllll}
\toprule
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Past Behaviour} & \multicolumn{4}{c}{Behavioural Intention} & \multicolumn{4}{c}{Action Probability} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13}
  & $\beta$ & b & SE & p & $\beta$ & b & SE & p & $\beta$ & b & SE & p\\
\midrule
Intercept & 0 & 0.1 & 0.04 & 0.0094 & 0 & 1.09 & 0.03 & <0.001 & 0 & 1.21 & 0.07 & <0.001\\
Guilt & 0.02 & 0.01 & 0.04 & 0.7562 & 0 & 0 & 0.03 & 0.942 & 0.05 & 0.06 & 0.06 & 0.326\\
Image Shame & -0.07 & -0.05 & 0.05 & 0.3926 & 0.1 & 0.06 & 0.04 & 0.154 & -0.02 & -0.02 & 0.1 & 0.826\\
Moral Shame & 0.48 & 0.31 & 0.06 & <0.001 & 0.56 & 0.33 & 0.05 & <0.001 & 0.59 & 0.77 & 0.11 & <0.001\\
\midrule
R$^{2}$ &  & 0.181 &  &  &  & 0.425 &  &  &  & 0.376 &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

Behavioral intentions and action probability correlated at \emph{r} = 0.86, \emph{p} \textless0.001, so we scaled and combined the two measures into a single indicator of future anti-war action. We ran a linear regression predicting future anti-war behavior from attitude towards the war, negative and positive emotions, and shame and guilt. Table \ref{tab:Table3} presents the results. Overall, the model explained 47\% of variance in future behavior. Anti-war behavioral intentions were best predicted by moral shame (\(\beta\) = 0.36, \emph{p} \textless0.001), followed by negative dominant emotions (\(\beta\) = 0.24, \emph{p} \textless0.001), and only then attitude (\(\beta\) = -0.15, \emph{p} \textless0.001). Negative submissive emotions had a small negative effect on intentions to act (\(\beta\) = -0.08, \emph{p} = 0.015).

\begin{table}[H]

\caption{\label{tab:Table3}Predicting future behavior from emotions and attitudes
}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llllll}
\toprule
  & $\beta$ & b & SE & t-value & p\\
\midrule
Intercept & 0 & -0.1 & 0.12 & -0.82 & 0.410\\
Emotions: Positive & -0.032 & -0.03 & 0.03 & -0.96 & 0.337\\
Emotions: Negative Dominant & 0.237 & 0.19 & 0.03 & 6.4 & <0.001\\
Emotions: Negative Submissive & -0.076 & -0.07 & 0.03 & -2.43 & 0.015\\
Attitudes towards the war & -0.152 & -0.12 & 0.03 & -3.77 & <0.001\\
\addlinespace
Group-based moral shame & 0.36 & 0.22 & 0.05 & 4.63 & <0.001\\
Group-based image shame & 0.002 & 0 & 0.04 & 0.03 & 0.977\\
Group-based guilt & 0.033 & 0.02 & 0.03 & 0.76 & 0.450\\
\bottomrule
\end{tabular}
\end{table}

The results of the zero-inflated negative binomial model (Table \ref{tab:TableS4}, SM) suggested that different predictors account for the presence vs.~absence of intentions to act versus the strength of these intentions. Only moral shame (\emph{b} = -.36, \emph{p} = .036) and attitude (\emph{b} = .38, \emph{p} = .001) predicted the probability of no action, whereas only negative emotions predicted the strength of these intentions. Negative dominant emotions predicted stronger intentions to act (\emph{b} = .24, \emph{p} \textless{} .001), whereas negative submissive emotions predicted weaker intentions to act (\emph{b} = -.14, \emph{p} = .013).

@@

\begin{table}[H]

\caption{\label{tab:TableS1}Path model predicting group-based guilt and shame from political beliefs.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llllll}
\toprule
  & $\beta$ & b & SE & t-value & p\\
\midrule
Intercept & 0 & -0.1 & 0.12 & -0.82 & 0.410\\
Emotions: Positive & -0.032 & -0.03 & 0.03 & -0.96 & 0.337\\
Emotions: Negative Dominant & 0.237 & 0.19 & 0.03 & 6.4 & <0.001\\
Emotions: Negative Submissive & -0.076 & -0.07 & 0.03 & -2.43 & 0.015\\
Attitudes towards the war & -0.152 & -0.12 & 0.03 & -3.77 & <0.001\\
\addlinespace
Group-based moral shame & 0.36 & 0.22 & 0.05 & 4.63 & <0.001\\
Group-based image shame & 0.002 & 0 & 0.04 & 0.03 & 0.977\\
Group-based guilt & 0.033 & 0.02 & 0.03 & 0.76 & 0.450\\
\bottomrule
\end{tabular}
\end{table}

The results of zero-inflated negative binomial regressions are presented in Table \ref{tab:TableS2}. This model allows us to test both what predicts the likelihood of experiencing these emotions (zero vs.~non-zero: the zero-inflation model) and what predicts the strength of these emotions for those who experience them (the count model). The results for strength of the emotions are identical to the results from the linear regression, but only democratic values predict the likelihood of experiencing these emotions in the first place.

\begin{table}[H]

\caption{\label{tab:TableS2}Zero-inflated negative binomial regression predicting group-based guilt and shame from political beliefs.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrrrrr}
\toprule
\multicolumn{1}{c}{\textbf{}} & \multicolumn{3}{c}{\textbf{Moral Shame}} & \multicolumn{3}{c}{\textbf{Image Shame}} & \multicolumn{3}{c}{\textbf{Guilt}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10}
\em{ } & \em{b} & \em{SE} & \em{p} & \em{b} & \em{SE} & \em{p} & \em{b} & \em{SE} & \em{p}\\
\midrule
\em{\textbf{Count model}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}}\\
\midrule
Intercept & -0.76 & 0.31 & 0.014 & -0.44 & 0.29 & 0.129 & -0.85 & 0.28 & 0.002\\
Democratic values & 0.23 & 0.05 & <0.001 & 0.17 & 0.05 & <0.001 & 0.25 & 0.05 & <0.001\\
Political cynicism (politicians) & 0.12 & 0.05 & 0.006 & 0.12 & 0.04 & 0.005 & 0.07 & 0.04 & 0.065\\
Political cynicism (politics) & -0.04 & 0.03 & 0.172 & -0.03 & 0.03 & 0.253 & -0.02 & 0.03 & 0.548\\
\addlinespace
Political alienation & -0.03 & 0.03 & 0.255 & <0.001 & 0.03 & 0.971 & <0.001 & 0.03 & 0.992\\
Political responsibility & 0.01 & 0.03 & 0.691 & <0.001 & 0.03 & 0.948 & 0.03 & 0.03 & 0.308\\
\midrule
\em{\textbf{Zero-inflation model}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}}\\
\midrule
Intercept & 5.03 & 0.63 & <0.001 & 5.15 & 0.62 & <0.001 & 4.17 & 0.6 & <0.001\\
Democratic values & -0.74 & 0.1 & <0.001 & -0.77 & 0.1 & <0.001 & -0.67 & 0.09 & <0.001\\
\addlinespace
Political cynicism (politicians) & -0.05 & 0.09 & 0.529 & -0.06 & 0.08 & 0.493 & 0.05 & 0.08 & 0.569\\
Political cynicism (politics) & -0.11 & 0.07 & 0.115 & -0.11 & 0.07 & 0.123 & -0.07 & 0.07 & 0.315\\
Political alienation & 0.02 & 0.07 & 0.73 & 0.04 & 0.07 & 0.579 & -0.06 & 0.07 & 0.369\\
Political responsibility & -0.05 & 0.07 & 0.528 & -0.05 & 0.07 & 0.492 & -0.07 & 0.07 & 0.295\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:TableS3} presents the results of zero-inflated negative binomial regression models predicting anti-war behavior from guilt and shame. Only moral shame consistently predicted anti-war action. Those who experienced more moral shame were more likely to have performed at least one action to oppose the war in the past (zero-inflation model) and were more willing and more certain that they will perform actions to oppose the war in the future (count model).

\begin{table}[H]
